---
title: Analyzing the 4K PBMC dataset
author: Aaron Lun and others
date: 28 December 2017
output: 
  BiocStyle::html_document:
    fig_caption: no
---

```{r, echo=FALSE, results="hide"}
library(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
opts_chunk$set(dpi=300, dev="png", dev.args=list(pointsize=15))
options(bitmapType="cairo")
```

# Introduction

This document performs a brief analysis of the 900 neuron dataset, with particular focus on the cells that are uniquely detected by EmptyDrops or CellRanger.
First, we load in the raw count matrix:

```{r}
library(DropletUtils)
fname <- "../../data/neurons_900/raw_gene_bc_matrices/mm10"
sce <- read10xCounts(fname, col.names=TRUE)
sce
```

... and define the cells with `emptyDrops`:

```{r}
set.seed(100)
e.out <- emptyDrops(counts(sce))
e.keep <- e.out$FDR <= 0.001
summary(e.keep)
```

... and CellRanger (with an expected 900 cells, as the name of the dataset suggests):

```{r}
c.keep <- defaultDrops(counts(sce), expected=900)
summary(c.keep)
```

We retain all cells that are detected by either EmptyDrops or CellRanger.
This makes it easier to compare the two methods later, rather than having to perform two separate analyses.

```{r}
keep <- c.keep | (e.keep & !is.na(e.keep))
detection <- rep("Both", length(keep))
detection[c.keep & !e.keep] <- "CellRanger"
detection[!c.keep & e.keep] <- "EmptyDrops"
table(detection)
```

Storing this in the metadata for future use.

```{r}
sce$Detection <- detection
sce$PValue <- e.out$PValue
sce <- sce[,keep]
```

# Adding gene-level annotation

We add some gene-level annotation.

```{r}
library(org.Mm.eg.db)
symb <- mapIds(org.Mm.eg.db, keys=rownames(sce), keytype="ENSEMBL", column="SYMBOL")
rowData(sce)$ENSEMBL <- rownames(sce)
rowData(sce)$SYMBOL <- symb
head(rowData(sce))
```

We relabel the rows with the gene symbols for easier reading.

```{r}
library(scater)
rownames(sce) <- uniquifyFeatureNames(rowData(sce)$ENSEMBL, rowData(sce)$SYMBOL)
head(rownames(sce))
```

We also determine the chromosomal location for each gene.

```{r}
library(TxDb.Mmusculus.UCSC.mm10.ensGene)
location <- mapIds(TxDb.Mmusculus.UCSC.mm10.ensGene, keys=rowData(sce)$ENSEMBL, 
    column="CDSCHROM", keytype="GENEID")
rowData(sce)$CHR <- location
summary(location=="chrM")
```

# Quality control on the cells

Cell detection can be considered an implicit quality control step, so technically, no further work is needed.
Nonetheless, we examine some commonly used metrics.

```{r qchist, fig.wide=TRUE, fig.asp=0.5}
sce <- calculateQCMetrics(sce, feature_controls=list(Mito=which(location=="chrM")))
par(mfrow=c(1,3))
hist(sce$log10_total_counts, breaks=20, col="grey80")
hist(sce$log10_total_features_by_counts, breaks=20, col="grey80")
hist(sce$pct_counts_Mito, breaks=20, col="grey80")
```

A large number of the features with low total counts have high mitochondrial proportions.
This is unsurprising and probably corresponds to damaged cells - we'll filter them out.

```{r mitoscatter, fig.wide=TRUE}
par(mfrow=c(1,2))
plot(sce$total_features_by_counts, sce$pct_counts_Mito)
plot(sce$total_counts, sce$pct_counts_Mito)
discard <- isOutlier(sce$pct_counts_Mito, nmads=3, type="higher")
sce <- sce[,!discard]
summary(discard)
```

# Examining gene expression

We have a look at the average expression of each gene.

```{r abhist}
ave <- calcAverage(sce)
rowData(sce)$AveCount <- ave
hist(log10(ave), col="grey80")
```

We also examine the top-most expressed genes.
This contains ribosomal protein genes and other usual suspects.

```{r highexpr, fig.wide=TRUE, fig.asp=1.5}
plotHighestExprs(sce)
```

# Normalizing for cell-specific biases

We cluster cells together to improve the accuracy of the pooling later.

```{r}
library(scran)
set.seed(1000)
clusters <- quickCluster(sce, pc.approx=TRUE)
table(clusters)
```

We then use the deconvolution method to compute size factors for each cell.

```{r}
sce <- computeSumFactors(sce, min.mean=0.1, cluster=clusters)
summary(sizeFactors(sce))
```

We can plot these against the library sizes to see how much of a difference it makes.

```{r sfplot}
plot(sce$total_counts, sizeFactors(sce), log="xy")
```

Finally, we compute normalized log-expresion values.

```{r}
sce <- normalize(sce)
```

# Modelling the mean-variance trend

We assume that the technical noise is Poisson and create a fitted trend on that basis.

```{r}
new.trend <- makeTechTrend(x=sce)
```

We actually estimate the variances and plot the trend against the original variances as well.

```{r trendplot}
fit <- trendVar(sce, use.spikes=FALSE, loess.args=list(span=0.05))
plot(fit$mean, fit$var, pch=16)
curve(fit$trend(x), col="dodgerblue", add=TRUE)
curve(new.trend(x), col="red", add=TRUE)
```

We decompose the variance and have a look at the genes with the highest residual.

```{r}
dec <- decomposeVar(fit=fit)
top.dec <- dec[order(dec$bio, decreasing=TRUE),] 
head(top.dec)
``` 

We can plot the genes with the largest biological components, to verify that they are indeed highly variable.

```{r hvgplot, fig.wide=TRUE}
plotExpression(sce, features=rownames(top.dec)[1:10])
```

# Dimensionality reduction

We use the `denoisePCA` function to perform PCA, using the assumed Poisson technical trend.

```{r}
sce <- denoisePCA(sce, technical=new.trend, approx=TRUE)
ncol(reducedDim(sce, "PCA"))
plot(attr(reducedDim(sce), "percentVar"))
```

We can plot the first few components.

```{r pcaplot}
plotPCA(sce, ncomponents=3, colour_by="Detection")
```

Same with using _t_-SNE for visualization.

```{r}
set.seed(1000)
sce <- runTSNE(sce, use_dimred="PCA", perplexity=30)
plotTSNE(sce, colour_by="Detection")
```

# Clustering with graph-based methods

We use the shared nearest neighbour method for clustering.

```{r}
snn.gr <- buildSNNGraph(sce, use.dimred="PCA")
clusters <- igraph::cluster_walktrap(snn.gr)
table(clusters$membership)
```

Plotting them out to verify separateness.

```{r}
sce$Cluster <- factor(clusters$membership)
plotTSNE(sce, colour_by="Cluster")
```

Also examining their modularity scores.
We look at the ratio of the observed and expected edge weights, as the raw modularity varies by orders of magnitudes across clusters.

```{r}
cluster.mod <- clusterModularity(snn.gr, sce$Cluster, get.values=TRUE)
log.ratio <- log2(cluster.mod$observed/cluster.mod$expected + 1)
library(pheatmap)
pheatmap(log.ratio, cluster_rows=FALSE, cluster_cols=FALSE, 
    color=colorRampPalette(c("white", "blue"))(100))
```

# Marker gene detection

Detecting marker genes for each cluster.

```{r}
marker.out <- findMarkers(sce, clusters=sce$Cluster, direction="up")
```

Having a look at how the clusters interact with the detection status, so we can focus on EmptyDrops-unique clusters.

```{r}
table(sce$Cluster, sce$Detection)
```

Focusing on two clusters which seem to be interneurons:

```{r heatmap, fig.wide=TRUE}
interneurons <- c("8", "10")
current <- marker.out[[interneurons[1]]]
chosen <- rownames(current)[current$Top <= 20]
plotHeatmap(sce, features=chosen, exprs_values="logcounts", 
    zlim=5, center=TRUE, symmetric=TRUE, cluster_cols=FALSE,
    colour_columns_by="Cluster", columns=order(sce$Cluster),
    show_colnames=FALSE)

current <- marker.out[[interneurons[2]]]
chosen <- rownames(current)[current$Top <= 20]
plotHeatmap(sce, features=chosen, exprs_values="logcounts", 
    zlim=5, center=TRUE, symmetric=TRUE, cluster_cols=FALSE,
    colour_columns_by="Cluster", columns=order(sce$Cluster),
    show_colnames=FALSE)
```

We'll record this in the metadata for future use.

```{r}
metadata(sce)$Interneurons <- interneurons
```

# Wrapping up

We save the various bits and pieces for further plotting.

```{r}
saveRDS(sce, file="sce.rds")
```

Printing the session information.

```{r}
sessionInfo()
```
